Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-03-15 ltx/BrainIDE.tex
some updates to the paper

For models to be clinically viable they must be patient specific. A possible approach to achieve this would be to use a general neural field model, like the Wilson and Cohen~\cite{Wilson1973} model or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995}, and fit it to a patients EEG data. Fitting the neural models to individuals is a highly non-trivial task, and until very recently this has not been reported in the literature. An estimation frame work for neural field model known as dynamical causal modeling (DCM)~\cite{David2003} \cite{David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set. The interconnected brain regions are modeled by the Jansen and Ritt equations. The approach has proven very useful in understanding specific hierarchical networks of neural information processing. Another recent publication describing a parameter estimation method with a neural field model used an unscented Kalman filter with the Wilson-Cowan neural field equations~\cite{schiff2008kalman}.  This work takes a more system theoretic approach to the neural estimation problem and marks a first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

\section{Neural Field Model}
In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behavior. It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.
The model relates the average number of action potentials $g$ arriving at $r$ to the local post synaptic membrane voltage. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by
\begin{equation}\label{SpikesToPotential}
v\left( {r,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {r,t'} \right)dt'} + e\left( {r,t} \right).
\end{equation}
The term $de(r,t)$ can be considered the potential generated by unmodeled inputs and spontaneous background activity and is a zero mean temporal Wiener process with incremental covariance $\Sigma_e(r)dt$. The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{SynapticRespKernel}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - \alpha t)} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.
\end{equation}
where $\alpha=\tau^{-1}$ and $\tau$ is the synaptic time constant. The synaptic response kernel can be more elaborate, using rise and fall time constants, however, this simple form serves our purposes. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{RateBasedInteractions}
g\left( r,t \right) = \int_\Omega  {w\left( r,r' \right)f\left( v\left( r',\bar t \right) \right)dr'}
\end{equation}
where $\bar t = t - d\left| {r - r'} \right|$, $d$ is the propagation delay of action potentials, $f$ is spiking rate of populations $r'$, $w$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface.

@@

Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( \frac{\partial }{dt} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + de\left( r,t \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
dv\left( {r,t} \right) + \alpha v\left( r,t \right)dt = \alpha\int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'dt}  +de\left( r,t \right).
\end{equation}

\section{Discrete Time and Space Model}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) = \alpha \left( -v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = \alpha q\left(v(r,t) \right) dt  + d e\left( r,t \right)
\end{equation}
A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \alpha \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + \alpha q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha \left( -v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \alpha \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  (1 - \delta \alpha)v(r_{ij},t_k) + \delta \alpha \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel. In Eq~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left()\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}.
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.

>>>>

For models to be clinically viable they must be patient specific. A possible approach to achieve this would be to use a general neural field model, like the Wilson and Cohen~\cite{Wilson1973} model or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995}, and fit it to a patients EEG data. Fitting the neural models to individuals is a highly non-trivial task, and until very recently this has not been reported in the literature. An estimation frame work for neural field models known as dynamical causal modeling (DCM)~\cite{David2003} \cite{David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set. The interconnected brain regions are modeled by the Jansen and Ritt equations. The approach has proven very useful in understanding specific hierarchical networks of neural information processing. Another recent publication describing a parameter estimation method with a neural field model used an unscented Kalman filter with the Wilson-Cowan neural field equations~\cite{schiff2008kalman}.  This work takes a more system theoretic approach to the neural estimation problem and marks a first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

\section{Neural Field Model}
In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.
The model relates the average number of action potentials $g(r,t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(r,t)$. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by
\begin{equation}\label{SpikesToPotential}
v\left( {r,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{SynapticRespKernel}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - \alpha t)} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.
\end{equation}
where $\alpha=\tau^{-1}$ and $\tau$ is the synaptic time constant. The synaptic response kernel can be more elaborate, using rise and fall time constants, however, this simple form serves our purposes. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{RateBasedInteractions}
g\left( r,t \right) = \int_\Omega  {w\left( r,r' \right)f\left( v\left( r',\bar t \right) \right)dr'} + e\left(r,t\right)
\end{equation}
where $\bar t = t - d\left| {r - r'} \right|$, $d$ is the propagation delay of action potentials, $f$ is spiking rate of populations $r'$, $w$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The term $de(r,t)$ can be considered the action potentials generated by unmodeled inputs and spontaneous background activity and is a Wiener process with mean $\mu$ incremental covariance $\Sigma_e(r)dt$.

@@

Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining Eqs.~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t h\left(t - t'\right) \left(\int\limits_\Omega  w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' + e\left(r,t\right)\right) dt'
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( \frac{\partial }{dt} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + de\left( r,t \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
dv\left( {r,t} \right) + \alpha v\left( r,t \right)dt = \alpha\int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'dt}  +de\left( r,t \right).
\end{equation}

\section{Discrete Time and Space Model}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) = \alpha \left( -v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = \alpha q\left(v(r,t) \right) dt  + d e\left( r,t \right)
\end{equation}
A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \alpha \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + \alpha q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta \alpha \left( -v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \alpha \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  (1 - \delta \alpha)v(r_{ij},t_k) + \delta \alpha \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel. In Eq~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}.
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.
